On the propriety of restricted Boltzmann machines

### 2016-10-22 writing/draft.Rmd
2nd edits from Nordman

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degeneracy, instability and/or uninterpretability in such models in the binary case. We also describe the difficulties that arise in likelihood-based fitting such highly flexible models, especially as Gibbs sampling methods are often advocated for the RBM model structure.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples]. A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data are generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. RBMs can often generate data lacking realistic variability and thereby fail to provide a satisfactory conceptualization of the data generation process. In addition to such degeneracy, we find that RBMs easily exhibit types of instability in the parameter space. In practice, this may seen when a single pixel change in an image results in a wildly different classification. Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep]. We investigate these phenomena in RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable. But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs. The literature often suggests MCMC approximations to fit these models (exploiting conditional structure in the hidden and visible variables), but little is said about monitoring convergence [hinton2010practical, @hinton2006fast]. Related to this, the MCMC algorithms involve updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods. In Section \ref{bayesian-model-fitting}, we compare three Bayesian modelling techniques (which can be computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of feasibility. Additionally, because these models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the empirical distribution of most realistic training sets. There is no "smoothing" achieved with a RBM fit using a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, Section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{bayesian-model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common. In this paper, we consider the binary case of an RBM for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

@@

### Joint distribution
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, as the potential encodings for a binary $H + V$-dimensional space, $\mathcal{X}$. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$containing two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion. A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of conditioal independence of the RBM (i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the visible and hidden layers.

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes. An exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, e.g. $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$, is close to the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows. When random variables $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, then the RBM model is near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$. Here the mean of $t(\boldsymbol x)$ isgiven as
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} t(\boldsymbol X)  \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability), it is useful to imagine how a data model (the RBM in particular) might be expanded to incorporate more and more observations. When increasing the size of a RBM model to handle more visible variables, it becomes necessary to expand the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables to be included). To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \theta_{VH}), V \ge 1$, denote a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ $(V \ge 1)$ as
\begin{align}
\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \{-1,1\}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.

Essentially, in formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$),the ratio (\ref{eqn:elpr}) should stay bounded, meaning that the largest probability under this model at parameter $\boldsymbol \theta_V$ should maintain a magnitude within the smallest probability allowed under the same model. This leads to the following definition:

\begin{defn}[unstable RBM]
Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if, as the number of visible variables increase ($V \rightarrow \infty$), it holds that 
\begin{align*}
\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}(\boldsymbol \theta_V) > C$ for all $V \ge V_C$.

Note that this definition of *S-unstable* is a generalization or re-interpretation of "unstable" used in @schweinberger2011instability allowing nonexponential family models for visibles $(v_1, \dots, v_V)$ and an increasing number of parameters. While this definition appears different, it matchs that of @schweinberger2011instability for the special exponential models considered there.

S-Unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \{-1,1\}^V \text{ differ by exactly one component}\right\},
$$
the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.

@@

Additionally, S-unstable RBM models are connected to degenerate models  placing all probability on a small portion of the sample space. To see this, define a modal set 
$$
M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$, and along a sequence $\boldsymbol \theta_V, V \ge 1$, of RBM parameters.

\begin{prop}
\label{prop:degen}
For an S-unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_V}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta_V}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
\end{prop}

In other words, as a consequence of S-unstable models, all probability in the RBM model formulation with a lerge number of visible variables will stack up on mode sets or simply those outcomes for visibles with with the highest probability. Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note. These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixed number of parameters (i.e. our results are broader for a more complicated model form for the visible variables).

>>>>

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, and two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based fitting of such highly flexible models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples]. A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, and with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal]. However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are often unsatisfactory as a conceptualization of how data are generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. At issue, RBMs can often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process. In addition to such degeneracy, we find that RBMs can easily exhibit types of instability in the parameter space. In practice, this may seen when a single pixel change in an image results in a wildly different classification. Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep]. We investigate and these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable. But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs. The literature often suggests Markov chain Monet Carlo (MCMC) approximations to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about monitoring convergence [@hinton2010practical, @hinton2006fast]. Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods. In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of feasibility. Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows. Section \ref{background} formally definesthe RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with proofs of main results.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common. In this paper, we consider the binary case of a RBM for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

@@

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_H, v_1,\dots,v_V)$ represent the states of the visible and hidden nodes in an RBM for some integers $V, H \ge 1$. We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, as the potential encodings for a binary $H + V$-dimensional space, $\mathcal{X}$. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function to ensure the probabilities (\ref{eqn:pmf}) sums to one. For  
\begin{align}
\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \{0,1\}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{X} = \{0,1\}^{V + H}, \label{eqn:t}
\end{align}
let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of random variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, V,\\j = 1, \dots, H}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion. A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture within the RBM (i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both the visible and hidden layers.

@@

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph. Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function for a valid distribution. Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$, is close to the boundary of $C$.

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows. When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \boldsymbol \tau \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X} = \{0,1\}^{V + H}\} \subset \{0,1\}^{V + H + V*H}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\boldsymbol \tau$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is given as
$$
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
$$

@@

Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*. Instability can be roughly characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability), it is useful to imagine how the RBM model might be expanded to incorporate more and more observations. When increasing the size of a RBM model to handle more visible variables, it becomes necessary to expand the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables to be included). To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
\begin{align}
\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
\end{align}
where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \{-1,1\}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.

In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will see that the ratio (\ref{eqn:elpr}) needs to stay bounded for the RBM model to be stable. This leads to the following definition:

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles. A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if, as the number of visible variables increase ($V \rightarrow \infty$), it holds that 
\begin{align*}
\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}(\boldsymbol \theta_V) > C$ for all $V \ge V_C$.

Note that this definition of *S-unstable* is a generalization or re-interpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models for visibles $(v_1, \dots, v_V)$ and an increasing number of parameters.

S-unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \{-1,1\}^V \text{ differ by exactly one component}\right\},
$$
denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.

@@

Additionally, S-unstable RBM models are connected to degenerate models  placing all probability on a small portion of the sample space. To see this, define a modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_V, V \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable models are guaranteed to be degenerate, as the following result shows.

\begin{prop}
\label{prop:degen}
For an S-unstable RBM model and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_V}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta_V}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
\end{prop}

In other words, in S-unstable RBM models with a large number of visible variables, all probability in the RBM model formulation  will stack up on those outcomes for visibles with with the highest probability. Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note. These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixed number of parameters.
